---
title: "EDA, group by and summarize"
author: "Rachel"
date: "10/10/2021"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6, 
  fig.asp = .6, 
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.colour = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


## Load the weather data 

```{r}
weather_df =  
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate(
    name = recode(
      id, 
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())
```

## `group by`

```{r}
weather_df %>% 
  group_by(name) 
```

## count some things 

Counting month observations 

```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n())

weather_df %>% 
  count(name)
```
we can use count 
```{r}
weather_df %>% 
  count(month, name = "n_obs")
```

**NEVER** USE BASE R's `TABLE`

```{r, eval = FALSE}
weather_df %>% 
  pull(month)
  table()
```

Other helpful counter
```{r}
weather_df %>% 
  group_by(month) %>% 
  summarize(n_obs = n(),
            n_days = n_distinct(date))
```

## A digression on 2x2 tables

```{r}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold", 
      tmax >= 5 ~ "not_cold",
      TRUE    ~""
    )
  ) %>% 
  group_by(name, cold) %>% 
  summarize(count = n())
```
```{r}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold", 
      tmax >= 5 ~ "not_cold",
      TRUE    ~""
    )
  ) %>% 
  janitor::tabyl(name, cold)
```

## General summaries 

It's nice that summarize produces a data frame

```{r}
weather_df %>% 
  group_by(name,month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prcp = mean(prcp, na.rm = TRUE), 
    medain_tmin = median(tmin, na.rm = TRUE)
  )
```

This is a dataframe! So we can make a plot. 

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    mean_prcp = mean(prcp, na.rm = TRUE), 
    medain_tmin = median(tmin, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() + 
  geom_line()
```

Suppose you want to summarize many columns. 

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(across(prcp:tmin, mean))
```

Reminder: sometimes your results are easier to read in another format. 

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(mean_tmax = mean(tmax)) %>% 
  pivot_wider(
    names_from = name, 
    values_from = mean_tmax
  ) %>% 
  knitr::kable(digits = 1)
```


## `group by` and `mutate`

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax
  ) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```

what about window functions 

ranking....

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  mutate(temp_rank = min_rank(tmax))
```

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    tmax_rank = min_rank(desc(tmax))
  ) %>% 
  filter(tmax_rank < 2)
```

lag

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(temp_change = tmax - lag(tmax)) %>% 
  summarize(
    temp_change_max = max(temp_change, na.rm = TRUE),
    temp_change_sd = sd(temp_change, na.rm = TRUE)
  )
```


lagged variables 
```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    lagged_tmax = lag(tmax, n = 1),
    tmax_diff = tmax - lagged_tmax
  ) %>% 
  summarize(diff_sd = sd(tmax_diff, na.rm = TRUE))
```

## Limitations 

What if my "summary" is a linear model...

well...

```{r, eval = FALSE}
weather_df %>% 
  group_by(name) %>% 
  summarize(cor_tmin_tmax = cor(tmin, tmax, use = "complete"))

weather_df %>% 
  filter(name == "CentralPark_NY") %>%
  lm(tmax ~ tmin, data = .)

weather_df %>% 
  group_by(name) %>% 
  summarize(lm = lm(tmax ~ tmin))
```






